### 2021-05-10 written/eLifeRevision/Response.tex

{\Large \textbf{Response to reviewers}} \\
(original comments in bold)

\textbf{Bittner and colleagues introduce a machine learning framework for maximum entropy inference of model parameter distributions that are consistent with certain emergent model properties, specified by the investigator. The approach is illustrated on several models of potential interest.}

\textbf{Reviewers were broadly enthusiastic about the potential usefulness of this methodology. However, the reviews and ensuing discussion revealed several points of concern about the manuscript and the approach. The full reviewer comments are included below.}

We thank the reviewers for the excellent and constructive feedback.

We have made significant changes to our writing and explanation of emergent property inference (EPI), done extensive comparisons of EPI to other parameter inference methods, and have seriously increased the quality and depth of our scientific analyses, which now yield strong theoretical results.  Our model analyses now focus on analyzing the structure of parameter distributions, which EPI captures using flexible approximations enabled by deep learning.  Here, we provide a list of key manuscript updates, and below we explain how these changes address each of the reviewer’s concerns.

List of key manuscript improvements:
\begin{itemize}
\item We completely overhauled our presentation of the parameter inference technique emergent property inference (EPI). See response to main concern \#1, especially noted by R1 and R3.
\item We added an entirely new results section to compare EPI to alternative parameter inference techniques (SMC-ABC and SNPE) as parameter count increases, which demonstrates the advantages of EPI in high dimensional parameter spaces. See response to main concern \#2, especially noted by R2 and R3.
\item EPI is used to understand how noise across neuron type populations governs excitatory variability in a model of primary visual cortex.  
Here, EPI yields visually striking, novel insight where the conventional analytic approach became infeasible with increasing neuron types.  See response to main concern \#4.
\item EPI is used to discover multiple parametric regimes of rapid task switching in a model of superior colliculus.  
These regimes are efficiently identified and mechanistically characterized using the probabilistic modeling tools afforded by EPI, and we relate the inferred connectivities to results from optogenetic silencing experiments in rats. See response to main concern \#4.
\end{itemize}

@@

\textbf{The main concerns are summarized as follows:}

\textbf{1. The methodology is not adequately explained. Both the body text and methods section present a somewhat selective description that is very hard to follow in places and should be checked and rewritten for clarity, completeness, notational consistency and correctness.}

Agreed.
In the introduction, we present our method by focusing on the key aspect of EPI that differentiates it from other approaches to inverse problems.
Specifically, we explain how current methodology infers the parameters producing computation by conditioning on \textit{exemplar datasets} (real or simulated), whereas in EPI we condition directly on the \textit{emergent properties} that define the computation.

\begin{displayquote}
Lines 46-54\\
``\quoting{intro_data_ep}''
\end{displayquote}

@@

We now frame the method within the more general context of parameter inference techniques in neuroscience, rather than the context of recent advancements in machine learning.
We have made concentrated efforts to simplify language and to relate to all relevant existing methodology.

\begin{displayquote}
Lines 55-65\\
``\quoting{intro_epi}''
\end{displayquote}

To improve clarity, we have overhauled the notation and presentation of emergent properties.
Emergent properties are now denoted with $\mathcal{X}$ to signify an implicit dataset predicated by the emergent property.
And rather than a vector of first- and second-moment constraints, we present the emergent property more readably (and equivalently) as mean and variance constraints on emergent property statistics:

\begin{displayquote}
Line 759 \\
``
\begin{equation} \tag{11}
\mathcal{X}: \mathbb{E}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\mu}, \text{Var}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\sigma}^2.
\end{equation}''
\end{displayquote}

Emergent properties and EPI are introduced and explained with this revised formalism in Section \ref{results_dgm}, and we have completely redone the Methods Section \ref{methods_EPI} to improve clarity, precisely explain EPI optimization (Section \ref{methods_AL_opt}), and derive equivalence to established bayesian inference techniques (Section \ref{methods_VI}).

@@

In our revised writing, we deemphasize the role of maximum entropy in the EPI algorithm, because this has largely served as a distraction in our experience.
We show in Section \ref{methods_VI} that EPI is a special case of variational inference, since maximum entropy is the same normative selection principle as variational bayesian methods.  
Therefore, it makes sense to present EPI in the main text as a statistical inference technique that constrains the predictions of the inferred parameters to be an emergent property, and leave the details of the maximum entropy to the technically proficient in Section \ref{methods_EPI}.

\begin{displayquote}
Lines 156-158\\
``\quoting{results_ME}''
\end{displayquote}

@@

Figure \ref{fig:STG} has been largely redone to reflect the modified presentation, and improve the pictorial representation of the method.
In Section \ref{methods_RNN}, we now show that EPI is the only simulation-based inference method that controls the predictions of its inferred distribution (Fig. \ref{fig:RNN}C-D).

Finally, we emphasize the utility of this deep inference technique for scientific inquiry in a new paragraph at the end of Section \ref{results_dgm}.

\begin{displayquote}
Lines 163-179\\
``\quoting{results_hess}''
\end{displayquote}

@@

In this analysis, we demonstrate the improved scalability of deep inference techniques (EPI and SNPE) with respect to the state-of-the-art approximate bayesian computation technique (SMC-ABC).
While controlling for architecture size (Fig. RNN1), we push the limits of SNPE through targeted hyperparameter modifications (Fig. RNN3), and show that EPI scales to higher dimensional RNN connectivities producing stable amplification than SNPE.
Furthermore, we emphasize that SNPE does not constrain the properties of the inferred parameters; many connectivities inferred by SNPE result in unstable or nonamplified models (Fig. 2C, Fig. RNN2).

\textbf{3. The extent to which this method is generally/straightforwardly applicable was in doubt. It seemed as though a significant amount of computation was required to do inference on one specified property and that the computation would need to be run afresh to query a new property. The methodology in the cosubmission (Macke) made clear that computation required for successive inferences is 'amortized' during training on random parameters. Moreover, EPI seemed less flexible than the cosubmission's approach in that it required a differentiable loss function. The complementarity and advantages of this approach as opposed to the cosubmission's approach are therefore unclear. }

The reviewers are right to point out these characteristics of EPI: it does not amortize across emergent properties, and it requires differentiability of the emergent properties of the model. 
Indeed, SNPE is more suitable for inference with nondifferentiable mechanistic models and scientific problems requiring many inferred parameter distributions.  
However, these relative drawbacks of EPI with respect to SNPE can be considered choices made in a trade-off between simulation-based inference approaches.

Unlike SNPE, EPI leverages gradients of the emergent property throughout optimization, which may lead to better efficiency and scalability (Section 3.3).  
The emergent properties of many models in neuroscience are tractably differentiable, four of which we analyze in this manuscript ranging across levels of biological realism, network size, and computational function.
This trade-off is now explained in the Discussion (Section \ref{discussion}).

\begin{displayquote}
Lines 439-451\\
``\quoting{discussion_gradients}''
\end{displayquote}

@@

Furthermore, EPI focuses the entire expressivity of the approximating deep probability distribution on a single distribution, rather than spreading this expressivity to some uncharacterized degree across the chosen training distribution of amortized posteriors in SNPE (Section 5.1.1 Related Approaches).

\begin{displayquote}
Lines 828-833\\
``\quoting{amortization}''
\end{displayquote}

@@

Finally, we emphasize that EPI does something fundamental that SNPE and other inference techniques cannot.  EPI learns parameter distributions whose predictions are constrained to produce the emergent property.  We show in Figure 2 and Supplementary Figure RNN2 that SNPE does not control the statistical properties of its predictions, resulting in the inference of many parameters that are not consistent with the desired emergent property.

\begin{displayquote}
Lines 232-242\\
``\quoting{rnn_constrained}''
\end{displayquote}

@@

\textbf{4. Some examples lack depth in their treatment (see reviewer comments) and in some cases the presentation is somewhat misleading. The STG example is not in fact a model of the STG. The cosubmission (Macke) uses a model close to the original Prinz et al model, which is a model of the pyloric subnetwork. It would be instructive to benchmark against this same model, including computation time/resources required. Secondly, the subsequent example (input-responsivity in a nonlinear sensory system) appeared to imply that EPI permits 'generation' and testing of hypotheses in a way that other methods do not. All the method really does is estimate a joint distribution of feasible parameters in a specific model which is manually inspected to propose hypotheses. Any other method (including brute force sampling) could be used in a similar way, so any claim that this is an integral advantage of EPI would be spurious. Indeed, one reviewer was confused about the origin of these hypotheses. While it is helpful to illustrate how EPI (and other methods) could be used, the writing needs to be far clearer in general and should clarify that EPI does not offer any new specific means of generating or evaluating hypotheses.}

We thank the reviewers for explaining how they found some of the
presentation misleading.  We have taken serious care in this manuscript to clarify a.)
what is novel, appreciable scientific insight provided by EPI, as well as b.) which scientific
analyses are made possible by EPI.

\begin{enumerate}[label=(\alph*)]
\item  In the revised manuscript we clarify that novel theoretical insights are not being made into the STG subcircuit model or the recurrent neural network models.  The STG subcircuit serves as a motivational example to explain how EPI works, and we use RNNs exhibiting stable amplification as a substrate for scalability analyses.

\begin{displayquote}
Lines 77-81 \\
``\quoting{intro_results1}''
\end{displayquote}

@@

We do produce strong theoretical insights into a model of primary visual cortex (Section 3.4) and superior colliculus (Section 3.5).  These analyses have substantially more depth than the previous manuscript.

\begin{displayquote}
Lines 81-89 \\
``\quoting{intro_results2}''
\end{displayquote}

\item The ability to infer a flexible approximation to a probability distribution constrained to produce an emergent property is novel in its own right (Figure 2).  The deep probability distribution fit by EPI facilitates the mode identification (via gradient ascent of the parameter log probability) and sensitivity measurement (via the measurement of the eigenvector of the Hessian at a parameter value). These mode identifications and sensitivity measurements are done in Sections 3.1 (Fig. 1E), 3.4 (Fig. 3E), and 3.5 (Fig. 4C).  By using this mode identification technique along the ridges of high parameter probability in the SC model, we identify the parameters transitioning between the two regimes.  Finally, the sensitivity dimensions of the SC model identified by EPI facilitated regime characterization through perturbation analyses (Fig. 4D).

@@

\end{enumerate}

Importantly, we do not claim that these theoretical insights were necessarily dependent on using the techniques in b.).  
One could have come to these conclusions via various combinations of techniques mentioned in Section 5.1.1 Related Methods.
In the case of the V1 model inference, the main point is to indicate that such insight can be afforded by EPI and its related methods, in contrast to the analytic derivations emblematic of practice in theoretical neuroscience.
\begin{displayquote}
Lines 301-309 \\
``\quoting{v1_summary}''
\end{displayquote}
In the case of the SC model inference, random sampling would have taken prohibitively long, and it is unclear how the continuum between the two connectivity regimes would have been identified with alternative techniques:
\begin{displayquote}
Lines 416-419 \\
``\quoting{sc_summary}''
\end{displayquote}

@@

As the reviewers indicate, the STG model analyzed in our manuscript is not that of Prinz et al. 2004, and thus not the model analyzed by the cosubmission.  We found the Prinz et al. model prohibitive to infer with EPI, since the gradients of spiking frequency from its simulation are quite computationally intensive.  However, we found it critical to show that it’s very possible to do inference in such biophysically realistic Morris-Lecar models when gradients are tractable, which is the case of the 5-neuron STG model we analyzed from Gutierrez et al. 2013.  This 5-neuron model represents the IC neuron (hub) and its coupling to the pyloric (fast) or gastric mill (slow) subcircuit rhythms.  In the introductory text, we refer to this model as the “STG subcircuit” model (rather than “STG model”), and we better clarify what aspect of the STG is being modeled in Results Section 3.1.

\begin{displayquote}
Lines 100-105 \\
``\quoting{stg_subcircuit}''
\end{displayquote}

The difference between this model and the STG model of the pyloric subnetwork is emphasized in Discussion:

% Within another quote so need to copy.
\begin{displayquote}
Lines 444-447 \\
``However, conditioning on the pyloric rhythm \cite{marder1992dynamic} in a model of the pyloric subnetwork model \cite{prinz2004similar} proved to be prohibitive with EPI.
The pyloric subnetwork requires many time steps for simulation and many key emergent property statistics (e.g. burst duration and phase gap) are not calculable or easily approximated with differentiable functions.\\
''
\end{displayquote}

@@

\textbf{5. There is a substantial literature on parameter sensitivity analysis and inference in systems biology, applied dynamical systems and control that has been neglected in this manuscript. The manuscript needs to acknowledge, draw parallels and explain distinctions from current methods (ABC, profile likelihood, deep learning approaches, gaussian processes, etc). The under-referencing of this literature deepened concerns about whether this approach represented an advance. DOIs for a small subset of potentially relevant papers include:}

\textbf{https://doi.org/10.1038/nprot.2014.025 \\
http://doi.org/10.1085/jgp.201311116 \\
http://doi.org/10.1016/j.tcs.2008.07.005 \\
http://doi.org/10.3182/20120711-3-BE-2027.00381 \\
http://doi.org/10.1093/bioinformatics/btm382 \\
http://doi.org/10.1111/j.1467-9868.2010.00765.x \\
http://doi.org/10.1098/rsfs.2011.0051 \\
https://doi.org/10.1098/rsfs.2011.0047 \\
http://doi.org/10.1214/16-BA1017 \\
https://doi.org/10.1039/C0MB00107D}

Thank you for pointing us to these references on sensitivity analyses and applied dynamical systems.  
We have incorporated many of them into the current manuscript and explain EPI’s relation to each class of techniques in Section 5.1.1 Related approaches.

\begin{displayquote}
Lines 849-862 \\
``\quoting{structural_id}''
\end{displayquote}

@@

\textbf{General Assessment: The authors introduce a machine learning framework for maximum entropy inference of model parameters which are consistent with certain emergent properties, which they call 'emergent property inference'. I think this is an interesting and direction, and this looks like a useful step towards this program. I think the paper could be improved with a more thorough discussion both of the broad principles their black box approach seeks to optimize, as well as the details of its implementation. I also think the detailed examples should be more self-contained. Finally I find this work to be somewhat misrepresented as a key to all of theoretical neuroscience. This approach may have some things to offer to the interesting problem of finding parameter regions of models, but this is not the entirety of, nor really a major part of theoretical neuroscience as I see it. }

We thank the reviewer for their positive comments and thoughtful feedback.
We have made serious effort to improve our presentation and explanation of EPI (see response to main concern \#1).
Furthermore, we have focused on clearly motivating and describing each neural circuit model studied in this manuscript.
Sufficient mathematical detail is written in each results section, while full details are presented in Methods.
All code for training EPI on these models and their analysis are available in well-documented scripts and notebooks in our github repository.

We modify our writing to clarify that EPI is not a key to all of theoretical neuroscience, but rather a powerful solution to inverse problems in neural circuit modeling.
Inverse problems are indeed a major part of theoretical neuroscience.
Their solution is necessary for the scientific evaluation of mathematical representations of neural circuits, and ultimately for progress in the field.
We clarify this point in the introduction:

\begin{displayquote}
Lines 28-37 \\
``\quoting{intro_first}''
\end{displayquote}

@@

\textbf{Other concerns: }

\textbf{(1) Maximizing the entropy of the distribution is not a reparameterization invariant exercise. That is, results depend on whether the model parameters contains rates or time constants, for example. I wonder if this approach is attempting to use a 'flat prior' in some sense which has the same reparameterization issue? Can the authors comment? }

The reviewer is correct to point out that maximum entropy solutions are not reparameterization invariant, and indeed the units matter.  The reviewer’s suggestion that the method is in some sense using a flat prior is also correct.  To clarify, EPI does not execute posterior inference, because there is no explicit dataset or specified prior belief in the EPI framework.  However, we derive the relation of EPI to bayesian inference in Section 5.1.6, which shows EPI uses a uniform prior when framed as variational inference.

\begin{displayquote}
Lines 1026-1029 \\
``\quoting{ME_prior}''
\end{displayquote}

@@

We appreciate this feedback.  We have adapted the introduction to clarify that we are focusing on solving inverse problems in theoretical neuroscience.

\textbf{(3) I am somewhat familiar with the stomatogastric circuit model, and so that is where I think I understand what they have done best. I don't understand what I should take away from their paper with regards to this model. Are there any findings that hadn't been appreciated before? What does this method tell us about the system and or its model? }

We clarify in point 4 above that we are not claiming to produce novel, appreciable scientific insight about the STG subcircuit model, which is used as a motivation example. The takeaway is that the conductance parameters producing intermediate hub frequency belong to a complex 2-D distribution, which EPI captures accurately, and that EPI can tell us the parameter changes away from the prototypical configuration that change hub frequency the most or least.  For example, for increases in $g_{\text{el}}$ and $g_{\text{synA}}$ according to the proportions of the degenerate dimension of parameter space, intermediate hub neuron frequency will be preserved in this model.  This suggests that in the real STG neural circuit, the IC neuron will remain at an intermediate frequency between the pyloric and gastric mill rhythms if parameter changes are made along such a dimension.

\begin{displayquote}
Lines 175 -177 \\
``The directionality of $\mathbf{v}_2$ suggests that changes in conductance along this parameter combination will most preserve hub neuron firing between the intrinsic rates of the pyloric and gastric mill rhythms.''
\end{displayquote}

@@

Another important question the reviewer brings up is how well one can analyze the high-dimensional parameter distributions that EPI produces?  
Indeed, these distributions become more challenging to understand and visualize in high dimensions.  
This is where the sensitivity measurements appearing in sections 3.1, 3.4, and 3.5 can be particularly useful.  
Even in high dimensions, trained deep probability distributions offer tractable quantitative assessments of how parametric combinations affect the emergent property that was conditioned upon.

\textbf{4. There are some very particular choices made in the applications and I wonder how general the conclusions are as a consequence. For example, in equation (5) the authors choose an arbitrary amount of variance 0.01$^2$ - why? In the same example, why look at y=0.1 and 0.5? }

In the current manuscript, we make sure to explain all choices of the emergent property constraints  Here, we show the description of each emergent property with equations omitted.

\begin{displayquote}
Section \ref{results_dgm}, Lines 139-143\\
``\quoting{stg_ep_text1}~\quoting{stg_ep_text2}.''
\end{displayquote}

### 2021-05-19 written/eLifeRevision/Bittner_eLife2020.tex
John's edits incorporated

\thispagestyle{plain}
{\Large Interrogating theoretical models of neural computation with deep inference} \\
Sean R. Bittner$^{1}$, Agostina Palmigiano$^{1}$, Alex T. Piet$^{2,3,4}$, Chunyu A. Duan$^{5}$, Carlos D. Brody$^{2,3,6}$, \\
Kenneth D. Miller$^{1}$, and John P. Cunningham$^{7}$.

@@

\linenumbers
\section{Abstract}
A cornerstone of theoretical neuroscience is the circuit model: a system of equations that captures a hypothesized neural mechanism.  
Such models are valuable when they give rise to an experimentally observed phenomenon -- whether behavioral or a pattern of neural activity -- and thus can offer insights into neural computation.
The operation of these circuits, like all models, critically depends on the choices of model parameters.
A key step is then to identify the model parameters consistent with observed phenomena: to solve the inverse problem.
To solve challenging inverse problems in neuroscience, statistical inference techniques have been used to infer parameters distributions most likely to have produced neural datasets.
In this work, we present a novel technique, emergent property inference (EPI), that brings the power and versatility of the modern probabilistic modeling toolkit to theoretical neuroscience.
When theorizing circuit models, scientists predominantly focus on reproducing computational properties rather than a particular dataset.
Our method uses deep neural networks to learn parameter distributions with complex structure that produce specific computational properties in circuit models.
This methodology is introduced through a motivational example inferring conductance parameters in a circuit model of the stomatogastric ganglion.
Then, with recurrent neural networks of increasing size, we show that EPI allows precise control over the behavior of inferred parameters, and that EPI scales better in parameter dimension than alternative techniques.
In the remainder of this work, we present novel theoretical findings gained through the examination of complex parametric structure captured by EPI.
In a model of primary visual cortex, we discovered how connectivity with multiple inhibitory subtypes shapes variability in the excitatory population.
Finally, in a model of superior colliculus, we identified and characterized two distinct regimes of connectivity that facilitate switching between opposite tasks amidst interleaved trials, mechanistically characterized each regime using probabilistic tools afforded by EPI, and found conditions where these circuit models reproduce results from optogenetic silencing experiments.
Beyond its scientific contribution, this work illustrates the variety of analyses possible once deep learning is harnessed towards solving theoretical inverse problems.

\section{Introduction}
\q{intro_first}{The fundamental practice of theoretical neuroscience is to use a mathematical model to understand neural computation, whether that computation enables perception, action, or some intermediate processing.  
A neural circuit is systematized with a set of equations -- the model -- and these equations are motivated by biophysics, neurophysiology, and other conceptual considerations \cite{kopell1988coupled,  marder1998biophysics, abbott2008theoretical, wang2010neurophysiological, o2015computational}.
The function of this system is governed by the choice of model \emph{parameters}, which when configured in a particular way, give rise to a measurable signature of a computation.   
The work of analyzing a model then requires solving the inverse problem: given a computation of interest, how can we reason about the space, manifold, or distribution of parameters that give rise to it?  
The inverse problem is crucial for reasoning about likely parameter values, uniquenesses and degeneracies, and predictions made by the model \cite{gutenkunst2007universally, erguler2011practical, mannakee2016sloppiness}.}

@@

\q{intro_data_ep}{Statistical inference, of course, requires quantification of the vague term \emph{computation}.  
In neuroscience, two perspectives are dominant.  
First, often we directly use an \emph{exemplar dataset}: a collection of samples that express the computation of interest, this data being gathered either experimentally in the lab or from a computer simulation.  
While in some sense the best choice given its connection to experiment \cite{paninski2018neural}, some drawbacks exist: these data are well known to have features irrelevant to the computation of interest \cite{niell2010modulation, saleem2013integration, musall2019single}, confounding inferences made on such data.  
Related to this point, use of a conventional dataset encourages conventional data likelihoods or loss functions, which focus on some global metric like squared error or marginal evidence, rather than the computation itself.}

@@

The parameter distributions producing a computation may be thin, curved, bent, or multimodal along various parameter axes and combinations.
It is by capturing and quantifying this complex structure that EPI offers scientific insight.
Traditional approximation families (e.g. mean-field or mixture of gaussians) are limited in the distributional structure they may learn.
To address such restrictions on expressivity, major advancements in machine learning have enabled the use of deep probability distributions as flexible approximating families for such complicated distributions \cite{rezende2015variational, papamakarios2019normalizing} (see Section \ref{methods_NF}).
However, the adaptation of deep probability distributions to the problem of theoretical circuit analysis requires recent developments in deep learning for  constrained optimization \cite{loaiza2017maximum}, and architectural  choices for efficient and expressive deep generative modeling \cite{dinh2017density, kingma2018glow}.
We detail our method, which we call emergent property inference (EPI) in Section \ref{results_dgm}.

Equipped with this method, we demonstrate the capabilities of EPI and present novel theoretical findings from its analysis.
\q{intro_results1}{First, we show EPI's ability to handle biologically realistic circuit models using a five-neuron model of the stomatogastric ganglion \cite{gutierrez2013multiple}: a neural circuit whose parametric degeneracy is closely studied \cite{goldman2001global}.
Then, we show EPI's scalability to high dimensional parameter distributions by inferring connectivities of recurrent neural networks (RNNs) that exhibit stable, yet amplified responses -- a hallmark of neural responses throughout the brain \cite{murphy2009balanced, hennequin2014optimal, bondanelli2019population}.}
\q{intro_results2}{In a model of primary visual cortex \cite{litwin2016inhibitory, palmigiano2020structure}, EPI reveals how the recurrent processing across different neuron-type populations shapes excitatory variability: a finding that we show is analytically intractable.
Finally, we investigated the possible connectivities of superior colliculus that allow execution of different tasks on interleaved trials \cite{duan2019collicular}.
EPI discovered a rich distribution containing two connectivity regimes with different solution classes.
We queried the deep probability distribution learned by EPI to produce a mechanistic understanding of neural responses in each regime. 
Intriguingly, the inferred connectivities of each regime reproduced results from optogenetic inactivation experiments in markedly different ways.}
These theoretical insights afforded by EPI illustrate the value of deep inference for the interrogation of neural circuit models.

@@

Second, once the model is selected, one must specify what the model should produce.
In this STG model, we are concerned with neural spiking frequency, which emerges from the dynamics of the circuit model (Fig. \ref{fig:STG}B).
An emergent property studied by Gutierrez et al. is the hub neuron firing at an intermediate frequency between the intrinsic spiking rates of the fast and slow populations.
This emergent property is shown in Figure \ref{fig:STG}C at an average frequency of 0.55Hz.
Our notion of intermediate hub frequency is not strictly 0.55Hz, but also moderate deviations of this frequency between the fast (.35Hz) and slow (.68Hz) frequencies.
%, which are quantified in the emergent property with variance {0.025$^2$Hz$^2$}.
%To continue our running STG example, one such emergent property is the phenomenon of \emph{network syncing} -- in certain parameter regimes, the frequency of the hub neuron matches that of the fast and slow populations at an intermediate frequency.

@@

\subsection{A deep generative modeling approach to emergent property inference} \label{results_dgm}
Emergent property inference (EPI) formalizes the three-step procedure of the previous section with deep probability distributions \cite{rezende2015variational, papamakarios2019normalizing}.
First, as is typical, we consider the model as a coupled set of noisy differential equations.  
In this STG example, the model activity (or state) $\mathbf{x} = \left[ x_{\text{f1}}, x_{\text{f2}}, x_{\text{hub}}, x_{\text{s1}}, x_{\text{s2}} \right]$ is the membrane potential for each neuron, which evolves according to the biophysical conductance-based equation:
\begin{equation} \label{eq:STG_model1}
\begin{split}
C_m \frac{d\mathbf{x}(t)}{dt} = -h(\mathbf{x}(t); \mathbf{z}) + d\mathbf{B}
 \end{split}
\end{equation} 
where $C_m$=1nF, and $\mathbf{h}$ is a sum of the leak, calcium, potassium, hyperpolarization, electrical, and synaptic currents, all of which have their own complicated dependence on activity $\mathbf{x}$ and parameters $\mathbf{z} = [g_{\text{el}}, g_{\text{synA}}]$, and $d\mathbf{B}$ is white gaussian noise \cite{gutierrez2013multiple} (see Section \ref{methods_stg_model} for more detail).

>>>>

\thispagestyle{plain}
{\Large Interrogating theoretical models of neural computation with emergent property inference} \\
Sean R. Bittner$^{1}$, Agostina Palmigiano$^{1}$, Alex T. Piet$^{2,3,4}$, Chunyu A. Duan$^{5}$, Carlos D. Brody$^{2,3,6}$, \\
Kenneth D. Miller$^{1}$, and John P. Cunningham$^{7}$.

@@

\linenumbers
\section{Abstract}
A cornerstone of theoretical neuroscience is the circuit model: a system of equations that captures a hypothesized neural mechanism.  
Such models are valuable when they give rise to an experimentally observed phenomenon -- whether behavioral or a pattern of neural activity -- and thus can offer insights into neural computation.
The operation of these circuits, like all models, critically depends on the choice of model parameters.
A key step is then to identify the model parameters consistent with observed phenomena: to solve the inverse problem.
In this work, we present a novel technique, emergent property inference (EPI), that brings the modern probabilistic modeling toolkit to theoretical neuroscience.
When theorizing circuit models, theoreticians predominantly focus on reproducing computational properties rather than a particular dataset.
Our method uses deep neural networks to learn parameter distributions with these computational properties.
This methodology is introduced through a motivational example inferring conductance parameters in a circuit model of the stomatogastric ganglion.
Then, with recurrent neural networks of increasing size, we show that EPI allows precise control over the behavior of inferred parameters, and that EPI scales better in parameter dimension than alternative techniques.
In the remainder of this work, we present novel theoretical findings gained through the examination of complex parametric structure captured by EPI.
In a model of primary visual cortex, we discovered how connectivity with multiple inhibitory subtypes shapes variability in the excitatory population.
Finally, in a model of superior colliculus, we identified and characterized two distinct regimes of connectivity that facilitate switching between opposite tasks amidst interleaved trials,  characterized each regime via insights afforded by EPI, and found conditions where these circuit models reproduce results from optogenetic silencing experiments.
Beyond its scientific contribution, this work illustrates the variety of analyses possible once deep learning is harnessed towards solving theoretical inverse problems.

\section{Introduction}
\q{intro_first}{The fundamental practice of theoretical neuroscience is to use a mathematical model to understand neural computation, whether that computation enables perception, action, or some intermediate processing.  
A neural circuit is systematized with a set of equations -- the model -- and these equations are motivated by biophysics, neurophysiology, and other conceptual considerations \cite{kopell1988coupled,  marder1998biophysics, abbott2008theoretical, wang2010neurophysiological, o2015computational}.
The function of this system is governed by the choice of model \emph{parameters}, which when configured in a particular way, give rise to a measurable signature of a computation.   
The work of analyzing a model then requires solving the inverse problem: given a computation of interest, how can we reason about the distribution of parameters that give rise to it?  
The inverse problem is crucial for reasoning about likely parameter values, uniquenesses and degeneracies, and predictions made by the model \cite{gutenkunst2007universally, erguler2011practical, mannakee2016sloppiness}.}

@@

\q{intro_data_ep}{Statistical inference, of course, requires quantification of the vague term \emph{computation}.  
In neuroscience, two perspectives are dominant.  
First, often we directly use an \emph{exemplar dataset}: a collection of samples that express the computation of interest, this data being gathered either experimentally in the lab or from a computer simulation.  
Though a natural choice given its connection to experiment \cite{paninski2018neural}, some drawbacks exist: these data are well known to have features irrelevant to the computation of interest \cite{niell2010modulation, saleem2013integration, musall2019single}, confounding inferences made on such data.  
Related to this point, use of a conventional dataset encourages conventional data likelihoods or loss functions, which focus on some global metric like squared error or marginal evidence, rather than the computation itself.}

@@

The parameter distributions producing a computation may be curved or multimodal along various parameter axes and combinations.
It is by quantifying this complex structure that EPI offers scientific insight.
Traditional approximation families (e.g. mean-field or mixture of gaussians) are limited in the distributional structure they may learn.
To address such restrictions on expressivity, advances in machine learning have used deep probability distributions as flexible approximating families for such complicated distributions \cite{rezende2015variational, papamakarios2019normalizing} (see Section \ref{methods_NF}).
However, the adaptation of deep probability distributions to the problem of theoretical circuit analysis requires recent developments in deep learning for  constrained optimization \cite{loaiza2017maximum}, and architectural  choices for efficient and expressive deep generative modeling \cite{dinh2017density, kingma2018glow}.
We detail our method, which we call emergent property inference (EPI) in Section \ref{results_dgm}.

Equipped with this method, we demonstrate the capabilities of EPI and present novel theoretical findings from its analysis.
\q{intro_results1}{First, we show EPI's ability to handle biologically realistic circuit models using a five-neuron model of the stomatogastric ganglion \cite{gutierrez2013multiple}: a neural circuit whose parametric degeneracy is closely studied \cite{goldman2001global}.
Then, we show EPI's scalability to high dimensional parameter distributions by inferring connectivities of recurrent neural networks (RNNs) that exhibit stable, yet amplified responses -- a hallmark of neural responses throughout the brain \cite{murphy2009balanced, hennequin2014optimal, bondanelli2019population}.}
\q{intro_results2}{In a model of primary visual cortex \cite{litwin2016inhibitory, palmigiano2020structure}, EPI reveals how the recurrent processing across different neuron-type populations shapes excitatory variability: a finding that we show is analytically intractable.
Finally, we investigated the possible connectivities of a superior colliculus model that allow execution of different tasks on interleaved trials \cite{duan2019collicular}.
EPI discovered a rich distribution containing two connectivity regimes with different solution classes.
We queried the deep probability distribution learned by EPI to produce a mechanistic understanding of neural responses in each regime. 
Intriguingly, the inferred connectivities of each regime reproduced results from optogenetic inactivation experiments in markedly different ways.}
These theoretical insights afforded by EPI illustrate the value of deep inference for the interrogation of neural circuit models.

@@

Second, once the model is selected, one must specify what the model should produce.
In this STG model, we are concerned with neural spiking frequency, which emerges from the dynamics of the circuit model (Fig. \ref{fig:STG}B).
An emergent property studied by Gutierrez et al. is the hub neuron firing at an intermediate frequency between the intrinsic spiking rates of the fast and slow populations.
This emergent property (EP) is shown in Figure \ref{fig:STG}C at an average frequency of 0.55Hz.
To be precise, we define intermediate hub frequency not strictly as 0.55Hz, but frequencies of moderate deviation from 0.55Hz between the fast (.35Hz) and slow (.68Hz) frequencies.
%, which are quantified in the emergent property with variance {0.025$^2$Hz$^2$}.
%To continue our running STG example, one such emergent property is the phenomenon of \emph{network syncing} -- in certain parameter regimes, the frequency of the hub neuron matches that of the fast and slow populations at an intermediate frequency.

@@

\subsection{Emergent property inference via deep generative models} \label{results_dgm}
Emergent property inference (EPI) formalizes the three-step procedure of the previous section with deep probability distributions \cite{rezende2015variational, papamakarios2019normalizing}.
First, as is typical, we consider the model as a coupled set of noisy differential equations.  
In this STG example, the model activity (or state) $\mathbf{x} = \left[ x_{\text{f1}}, x_{\text{f2}}, x_{\text{hub}}, x_{\text{s1}}, x_{\text{s2}} \right]$ is the membrane potential for each neuron, which evolves according to the biophysical conductance-based equation:
\begin{equation} \label{eq:STG_model1}
\begin{split}
C_m \frac{d\mathbf{x}(t)}{dt} = -h(\mathbf{x}(t); \mathbf{z}) + d\mathbf{B}
 \end{split}
\end{equation} 
where $C_m$=1nF, and $\mathbf{h}$ is a sum of the leak, calcium, potassium, hyperpolarization, electrical, and synaptic currents, all of which have their own complicated dependence on activity $\mathbf{x}$ and parameters $\mathbf{z} = [g_{\text{el}}, g_{\text{synA}}]$, and $d\mathbf{B}$ is white gaussian noise \cite{gutierrez2013multiple} (see Section \ref{methods_stg_model} for more detail).

Second, we determine that our model should produce the emergent property of ``intermediate hub frequency" (Figure \ref{fig:STG}C).
\q{stg_ep_text1}{We stipulate that the hub neuron's spiking frequency -- denoted by statistic $\omega_{\text{hub}}(\mathbf{x})$ -- is close to a frequency of 0.55Hz, between that of the slow and fast frequencies.
Mathematically, we define this emergent property with two constraints: that the mean hub frequency is 0.55Hz,}
\begin{equation}\label{eq:EP_STG1}
 \mathbb{E}_{\mathbf{z},\mathbf{x}}\begin{bmatrix} \omega_{\text{hub}}(\mathbf{x}; \mathbf{z}) \end{bmatrix}  ~~=~~  0.55
\end{equation}
\q{stg_ep_text2}{and that the variance of the hub frequency is moderate}
\begin{equation}\label{eq:EP_STG2}
\text{Var}_{\mathbf{z},\mathbf{x}}\begin{bmatrix} \omega_{\text{hub}}(\mathbf{x}; \mathbf{z}) \end{bmatrix}  ~~=~~   0.025^2
\end{equation}
In the emergent property of intermediate hub frequency, the statistic of hub neuron frequency is an expectation over the distribution of parameters $\mathbf{z}$ and the distribution of the data $\mathbf{x}$ that those parameters produce.
We define the EP as the collection of these two constraints.
In general, an emergent property is a collection of constraints on statistical moments that together define the computation of interest.
